A Short and Mechanized Consistency Proof for Dependent Type Theory

### 2024-01-08 paper/paper.tex
Finish the fundamental theorem

The key technique that underlies our consistency proof is proof by
logical relation. In
Section~\ref{sec:spec}, we present the dependent type theory of
interest. In Section~\ref{sec:logreldep}, we give the definition
of the logical relation for the dependent type theory. Rather than
presenting the logical relation as an inductive-recursive definition,
we use the more elementary concept of a partial function to capture
the interpretation of types. The alternative representation requires us
to show that the set of equations indeed defines a partial function;
that is, for each input, there should always be a unique
output.
From the interpretation function, we can define the semantic
typing judgment for the set of lambda terms.
In Section~\ref{sec:logrelproof}, we prove the fundamental lemma,
which states that syntactic typing implies semantic typing. Once the
fundamental lemma is established, logical consistency follows as a
trivial corollary.

The downside of a pen and paper proof is that the validity of the
definitions themselves can be dubious at times. When working with a
dependent type theory that has a universe hierarchy, an intuitive
understanding of naive set theory no longer suffices as the size of
sets matters. For example, the partial function representing the
interpretation of types does not immediately correspond to a valid
definition in a proof assistant. We cannot easily define it as a
function as proof assistants such as Coq and Agda only accept total
functions. We cannot naively encode the interpretation function as an
inductive relation either since we violate the strict positivity
constraint. A detailed discussion of how we work around such technical
issues can be distracting. However, what can be equally distracting is
following along a proof where even the validity of the statement is
not trusted. As a result, we clearly mark the ``suspicious''
definitions, and defer the discussion of these definitions
until Section~\ref{sec:logrelmech}, where we give an overview of our
short Coq mechanization proof and the various proof engineering
techniques we adopt to make the mechanization for tractable.

Finally, in Section~\ref{sec:relatedwork}, we give a short survey of
existing literature related to logical consistency about dependent
type theory.

@@

\begin{figure}[h]
\[
\begin{array}{lcll}
\mathit{Natural\ numbers}\\
[[i]],[[j]],[[n]] & ::= & [[0]]\ |\ [[S i]] & \mbox{zero, successor} \\ \\

\mathit{Contexts}\\
[[G]]       & ::= & [[empty]]\ |\ [[G ++ A]] &  \\ \\
\mathit{Terms}\\
[[a]],[[b]],[[c]],[[t]],[[p]],[[A]],[[B]] & ::= & [[Set i]]\ |\ [[i]]\  |\ [[Void]]
                  & \mbox{universes, variables, empty type} \\
            & |   & [[Pi A B]]\ |\ [[\ A a]]\ |\ [[a b]]
                  & \mbox{function types, abstractions, applications} \\
            & |   & [[a ~ b : A ]]\ |\  [[refl]]\ |\ [[J t a b p]]
                  & \mbox{equality types, reflexivity proof, J eliminator} \\
            & |   & [[Bool]]\ |\  [[true]]\ |\  [[false]]\ |\  [[if a b0 b1]]
                  & \mbox{boolean type, true, false, if}
\end{array}
\]
  \caption{Syntax of \lang}
  \label{fig:syntax}
\end{figure}

@@

@@

\begin{lemma}[Pi Inv Alt]
  \label{lemma:piinvalt}
  Suppose $[[Interp I i Pi A B S]]$, then there exists some $[[S0]]$
  such that the following constraints hold:
  \begin{itemize}
  \item $[[Interp I i A S0]]$
  \item $[[forall a, (# a in S0 implies (# exists S1 , Interp I i B {a}
    S1 #) #)]]$
  \item $[[S = { b | forall a, (# a in S0 , forall
    S1, (# Interp I i B {a} S1,  b a in F a #) #) }]]$
  \end{itemize}
\end{lemma}
\begin{proof}
  Immedaite from Lemmas~\ref{lemma:interpinv} and \ref{lemma:logreldeter}.
\end{proof}

@@

\section{Semantic Typing and Consistency}
\label{sec:logrelproof}
\begin{figure}[h]
  \begin{equation*}
    \begin{split}
      [[rho |= G]] &:= \forall i\ j\ S, \text{ if }[[i < |G|]]\text{ and
                     } [[InterpR j (up S i G i) { rho } S ]] \text{, then } [[rho i in S]] \\
      [[G |= a : A]] &:= \forall [[rho]], \text{ if }[[rho |=
                       G]]\text{ then there exists some } [[j]] \text{
                       and } [[S]] \text{ such that } [[InterpR j A
                       {rho} S]] \text{ and } [[a {rho} in S]] \\
      [[|= G]] &:= \forall [[i < |G|]], \exists [[j]], [[drop S i G |= G i : Set j]]
    \end{split}
  \end{equation*}
  \caption{Semantic Typing for \lang}
  \label{fig:semtyping}
\end{figure}

@@

Next, we show that semantic typing is implied by syntactic typing
case-by-case through a series of lemmas.

\begin{lemma}[Set Inv]
  \label{lemma:setinv}
  The following two statements are equivalent:
  \begin{itemize}
  \item $[[G |= A : Set i]]$
  \item For all $[[rho]]$, if $[[rho |= G]]$, then there exists
    $[[S]]$ such that $[[InterpR i (A {rho}) S]]$
  \end{itemize}
\end{lemma}
\begin{proof}
  The forward direction is immediate by
  Lemma~\ref{lemma:interpinv}. We now consider the backward direction
  and show that $[[G |= A : Set i]]$ given the second bullet.

@@

The semantic context well-formedness judgment ($[[|= G]]$), unlike its syntactic
counterpart $[[|- G]]$, is defined through a for all quantified statement rather
than inductively over the context. It is easy to recover the same
structural rules:
\begin{lemma}[TODO]

Likewise, the semantic well-formendness judgment for substitutions
satisfies similar structural rules.
\begin{lemma}[Well-formed $[[rho]]$ cons]
  If $[[InterpR i A S]]$, $[[a in S]]$, and $[[rho |= G]]$, then
  $[[rho .: a |= G ++ A]]$.
\end{lemma}

\begin{lemma}[ST-Var]
  If $[[|= G]]$ and $[[i < |G|]]$, then $[[G |= i : up S i G i]]$.
\end{lemma}

\begin{lemma}[ST-Set]
  If $[[i < j]]$, then $[[G |= Set i : Set j]]$.
\end{lemma}
\begin{proof}
  Immedaite by Lemma~\ref{lemma:setinv} and \rref{I-Set}.
\end{proof}

\begin{lemma}[ST-Pi]
  If $[[G |= A : Set i]]$ and $[[G ++ A |= B : Set i]]$, then $[[G |= Pi
  A B : Set i]]$.
\end{lemma}
\begin{proof}
  Applying Lemma~\ref{lemma:setinv} to the
  conclusion, it now suffices to show that given $[[rho |= G]]$, there
  exists some $[[S]]$ such that $[[InterpR i Pi A B S]]$.
  From Lemma~\ref{lemma:setinv} and $[[G |= A : Set i]]$, we know that
  there exists some set $[[S0]]$ such that $[[InterpR i A S0]]$.
From $[[G ++ A |= B : Set i]]$, we know that there must
exists $[[S]]$ such that $[[InterpR i B {rho .: a} S]]$ for every $[[a
in S0]]$. The conclusion immediately follows from Lemma~\ref{lemma:piintroalt}.
\end{proof}

>>>>

The key technique that underlies our consistency proof is proof by
logical relation. In
Section~\ref{sec:spec}, we present the dependent type theory of
interest. In Section~\ref{sec:logreldep}, we give the definition
of the logical relation for the dependent type theory. Rather than
presenting the logical relation as an inductive-recursive definition,
we use the more elementary concept of a partial function to capture
the interpretation of types. The alternative representation requires us
to show that the set of equations indeed defines a partial function;
that is, for each input, there should always be a unique
output.
From the interpretation function, we can define the semantic
typing judgment for the set of lambda terms.
In Section~\ref{sec:logrelproof}, we prove the fundamental lemma,
which states that syntactic typing implies semantic typing. Once the
fundamental lemma is established, logical consistency follows as a
trivial corollary. In Section~\ref{sec:logrelmech}, we point out the
specifics related to the Coq mechanization of the proof described in earlier
section.
Finally, in Section~\ref{sec:relatedwork}, we give a short survey of
existing literature related to logical consistency about dependent
type theory.

@@

\begin{figure}[h]
\[
\begin{array}{lcll}
\mathit{Natural\ numbers}\\
[[i]],[[j]],[[n]] & \in &  [[SNat]] &  \\ \\

\mathit{Contexts}\\
[[G]]       & ::= & [[empty]]\ |\ [[G ++ A]] &  \\ \\
\mathit{Terms}\\
[[a]],[[b]],[[c]],[[t]],[[p]],[[A]],[[B]] & ::= & [[Set i]]\ |\ [[i]]\  |\ [[Void]]
                  & \mbox{universes, variables, empty type} \\
            & |   & [[Pi A B]]\ |\ [[\ A a]]\ |\ [[a b]]
                  & \mbox{function types, abstractions, applications} \\
            & |   & [[a ~ b : A ]]\ |\  [[refl]]\ |\ [[J t a b p]]
                  & \mbox{equality types, reflexivity proof, J eliminator} \\
            & |   & [[Bool]]\ |\  [[true]]\ |\  [[false]]\ |\  [[if a b0 b1]]
                  & \mbox{boolean type, true, false, if} \\ \\
\mathit{Substitution}\\
[[rho]] & \in & [[SNat -> STm]] &
\end{array}
\]
  \caption{Syntax of \lang}
  \label{fig:syntax}
\end{figure}

@@

The reflexivity of parallel reduction enables us to embed rules from
call-by-name semantics into parallel reduction, as the following lemma
shows.
\begin{lemma}[Par AbsCbn] For all $[[A]], [[a]],$ and $[[b]]$,
  \label{lemma:parabscbn}
  $[[(\ A a) b => b {a}]]$
\end{lemma}
\begin{proof}
  Immediate from Lemma~\ref{lemma:parrefl} and \rref{P-AppAbs}.
\end{proof}

@@

\begin{lemma}[Pi Inv Alt]
  \label{lemma:piinvalt}
  Suppose $[[Interp I i Pi A B S]]$, then there exists some $[[S0]]$
  such that the following constraints hold:
  \begin{itemize}
  \item $[[Interp I i A S0]]$
  \item $[[forall a, (# a in S0 implies (# exists S1 , Interp I i B {a}
    S1 #) #)]]$
  \item $[[S = { b | forall a, (# a in S0 , forall
    S1, (# Interp I i B {a} S1,  b a in S1 #) #) }]]$
  \end{itemize}
\end{lemma}
\begin{proof}
  Immedaite from Lemmas~\ref{lemma:interpinv} and \ref{lemma:logreldeter}.
\end{proof}

@@

\section{Semantic Typing and Consistency}
\label{sec:logrelproof}
\begin{figure}[h]
  \begin{equation*}
    \begin{split}
      [[rho |= G]] &:= \forall i\ j\ S, \text{ if }[[i < |G|]]\text{ and
                     } [[InterpR j (up Suc i G i) { rho } S ]] \text{, then } [[rho i in S]] \\
      [[G |= a : A]] &:= \forall [[rho]], \text{ if }[[rho |=
                       G]]\text{ then there exists some } [[j]] \text{
                       and } [[S]] \text{ such that } [[InterpR j A
                       {rho} S]] \text{ and } [[a {rho} in S]] \\
      [[|= G]] &:= \forall [[i < |G|]], \exists [[j]], [[drop Suc i G |= G i : Set j]]
    \end{split}
  \end{equation*}
  \caption{Semantic Typing for \lang}
  \label{fig:semtyping}
\end{figure}

@@

The following lemma makes the statement $[[G |= A : Set i]]$ easier to
work with.
\begin{lemma}[Set Inv]
  \label{lemma:setinv}
  The following two statements are equivalent:
  \begin{itemize}
  \item $[[G |= A : Set i]]$
  \item For all $[[rho]]$, if $[[rho |= G]]$, then there exists
    $[[S]]$ such that $[[InterpR i (A {rho}) S]]$
  \end{itemize}
\end{lemma}
\begin{proof}
  The forward direction is immediate by
  Lemma~\ref{lemma:interpinv}. We now consider the backward direction
  and show that $[[G |= A : Set i]]$ given the second bullet.

@@

The semantic context well-formedness judgment ($[[|= G]]$), unlike its syntactic
counterpart $[[|- G]]$, is defined through a for all quantified statement rather
than inductively over the context. It is easy to recover the same
structural rules:
\begin{lemma}[Semantic context well-formedness cons]
  \label{lemma:semwffcons}
  If $[[|= G]]$ and $[[G |= A : Set i]]$, then $[[|= G ++ A]]$.
\end{lemma}
\begin{proof}
  By the definition of semantic context well-formedness, the goal is
  to show that given $[[i < | G ++ A |]] = [[Suc |G|]]$, $[[drop Suc i
  (G ++ A) |= (G ++ A) i : Set j]]$. The statement can be easily
  proven by case analysis on whether $[[i]]$ is zero.
\end{proof}

Likewise, the semantic well-formendness judgment for substitutions
satisfies similar structural rules.
\begin{lemma}[Well-formed $[[rho]]$ cons]
  If $[[InterpR i A S]]$, $[[a in S]]$, and $[[rho |= G]]$, then
  $[[rho .: a |= G ++ A]]$.
\end{lemma}
\begin{proof}
  Start by unfolding the definition of $[[rho .: a |= G ++ A]]$ and
  performing a case analysis similar to the proof of
  Lemma~\ref{lemma:semwffcons}.  The case where the number is $[[0]]$
  requires Lemma~\ref{lemma:logreldeterhet} to finish the proof.
\end{proof}

Next, we show some non-trivial cases of the fundamental theorem as
top-level lemmas and leave the remaining cases as exercises for the reader.
\begin{lemma}[ST-Var]
  \label{lemma:stvar}
  If $[[|= G]]$ and $[[i < |G|]]$, then $[[G |= i : up Suc i G i]]$.
\end{lemma}
\begin{proof}
  TODO: formulate renaming
\end{proof}

\begin{lemma}[ST-Set]
  \label{lemma:stset}
  If $[[i < j]]$, then $[[G |= Set i : Set j]]$.
\end{lemma}
\begin{proof}
  Immedaite by Lemma~\ref{lemma:setinv} and \rref{I-Set}.
\end{proof}

\begin{lemma}[ST-Pi]
  \label{lemma:stpi}
  If $[[G |= A : Set i]]$ and $[[G ++ A |= B : Set i]]$, then $[[G |= Pi
  A B : Set i]]$.
\end{lemma}
\begin{proof}
  Applying Lemma~\ref{lemma:setinv} to the
  conclusion, it now suffices to show that given $[[rho |= G]]$, there
  exists some $[[S]]$ such that $[[InterpR i Pi A{rho} B{up rho} S]]$.
  From Lemma~\ref{lemma:setinv} and $[[G |= A : Set i]]$, we know that
  there exists some set $[[S0]]$ such that $[[InterpR i A {rho} S0]]$.
From $[[G ++ A |= B : Set i]]$, we know that there must
exists $[[S]]$ such that $[[InterpR i B {rho .: a} S]]$ for every $[[a
in S0]]$. The conclusion immediately follows from Lemma~\ref{lemma:piintroalt}.
\end{proof}

\begin{lemma}[ST-Abs]
  \label{lemma:stabs}
  If $[[G |= Pi A B : Set i]]$ and $[[G ++ A |= b : B]]$, then $[[G |=
  \ A b : Pi A B]]$.
\end{lemma}
\begin{proof}
  By unfolding the definition of $[[G |= \ A b : Pi A B]]$, we need to
  show that given some $[[rho |= G]]$, there exists some $[[i]]$ and
  $[[S]]$ such that $[[InterpR i Pi A {rho} B {up rho} S]]$ and $[[\
  A{rho} b{up rho} in S]]$.

  By Lemma~\ref{lemma:setinv} and the premise $[[G |= Pi A B : Set
  i]]$, there exists some set $[[S]]$ such that
  $[[InterpR i Pi A {rho} B {up rho} S]]$. It now suffices to show that
  $[[\A{rho} b{up rho} in S
  ]]$. By Lemma~\ref{lemma:piinvalt}, there exists some $[[S0]]$ such
  that all following conditions hold:
  \begin{itemize}
  \item $[[InterpR i A{rho} S0]]$
  \item $[[forall a, (# a in S0 implies (# exists S1 , InterpR i B
    {rho .: a}
    S1 #) #)]]$
  \item $[[S = { b | forall a, (# a in S0 , forall
      S1, (# InterpR i B {rho .: a} S1,  b a in S1 #) #) }]]$
  \end{itemize}
  To show that $[[\A{rho} b{up rho} in S]]$, we need to prove
  that given $[[a in S0]]$,
  $[[Interp I i B {rho .: a} S1]]$, we have  $[[( \A{rho} b{up rho} )
  a in S1]]$.
  By Lemma~\ref{lemma:logrelbackclos}, the set $[[S1]]$ is closed
  under expansion. By Lemma~\ref{lemma:parabscbn}, since $[[( \A{rho} b{up rho} )
  a => b {up rho} {a}]] = [[b {rho .: a}]]$, it suffices to show that
  $[[b {rho .: a} in S1]]$, which is immediate from $[[G ++ A |= b :
  B]]$ and the fact that the logical relation is deterministic and
  cumulative (Lemma~\ref{lemma:logreldeterhet}).
\end{proof}

\begin{lemma}[ST-App]
  \label{lemma:stapp}
  If $[[G |= b : Pi A B]]$ and $[[G |= a : A]]$, then $[[G |= b a : B {a}]]$.
\end{lemma}
\begin{proof}
Suppose $[[rho |= G]]$. The goal is to show that there exists some
$[[i]]$ and $[[S1]]$
such that  $[[b {rho} a {rho} in S1 ]]$ and $[[InterpR i B {a} {rho}
S1]]$, or equivalently, $[[InterpR i B {rho .: a {rho}} S1]]$ since
$[[B {a}{rho}]] = [[B {rho .: a {rho}}]]$. By the premise $[[G |= b :
Pi A B]]$, Lemma~\ref{lemma:setinv}, and Lemma~\ref{lemma:piinvalt},
there exists some $[[i]]$ and $[[S0]]$ such that:
  \begin{itemize}
  \item $[[InterpR i A{rho} S0]]$
  \item $[[forall a0, (# a0 in S0 implies (# exists S1 , InterpR i B
    {rho .: a0}
    S1 #) #)]]$
  \item $[[forall a0, (# a0 in S0 , forall
      S1, (# InterpR i B {rho .: a0} S1,  b {rho} a0 in S1 #) #)]]$
  \end{itemize}
  Instantiating the variable $[[a0]]$ from the last two bullets with
  the term $[[a {rho}]]$, the conclusion immediately follows.
\end{proof}

\begin{theorem}[The Fundamental Theorem]
  \label{theorem:soundness}
  If $[[G |- a : A]]$, then $[[G |= a : A]]$.
\end{theorem}
\begin{proof}
  Proof by structural induction over the derivation of $[[G |- a :
  A]]$. Lemmas~\ref{lemma:stvar},~\ref{lemma:stset},~\ref{lemma:stpi},~\ref{lemma:stabs},~\ref{lemma:stapp}
  can be used to discharge their syntactic counterpart
  (e.g. Lemma~\ref{lemma:stabs} for case \rref{T-Abs}). The remaining
  cases not covered by the lemmas are similar to the ones already
shown or simpler and therefore omitted from the text.
\end{proof}
